---
output: github_document
params:
   cplex_dir: !r Sys.getenv("CPLEX_HOME")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![Build
Status](https://travis-ci.org/ctlab/mwcsr.svg?branch=master)](https://travis-ci.org/ctlab/mwcsr) [![codecov](https://codecov.io/github/ctlab/mwcsr/branch/master/graphs/badge.svg)](https://codecov.io/gh/ctlab/mwcsr/branch/master) 

# mwcsr 

A package for solving maximum weight connected subgraph (MWCS) problem and its variants.

Supported MWCS variants are:

* classic (simple) MWCS, where only vertices are weighted;
* budget MWCS, where vertices are parametrized by costs and overall budget is limited; 
* generalized MWCS (GMWCS), where both vertices and edges are weighted;
* signal generalized MWCS (SGMWCS), where both vertices and edges are marked with weighted "signals", and a weight of a subgraph is calculated as a sum of weights of its unique signals.

Currently, four solvers are supported:

* heuristic relax-and-cut solver `rmwcs_solver` for MWCS and Budget MWCS;
* heuristic relax-and-cut solver `rnc_solver` for MWCS/GMWCS/SGMWCS;
* heuristic simulated annealing solver `annealing_solver` for MWCS/GMWCS/SGMWCS;
* exact (if CPLEX library is available) or heuristic (without CPLEX) solver `virgo_solver` 
for MWCS/GMWCS/SGMWCS. 


## Installation

The package can be installed from GitHub using `devtools`:

```{r eval=F}
library(devtools)
install_github("ctlab/mwcsr")
```

## Quick start

Load `mwcsr`, as well as `igraph` package, which contains functions for graph manipulations.

```{r message=FALSE}
library(mwcsr)
library(igraph)
```

Let's load an example instance of MWCS problem. The instance is a simple `igraph` object with `weight` vertex attribute.

```{r}
data("mwcs_example")
print(mwcs_example)
summary(V(mwcs_example)$weight)
```

Now let us initialize a heuristic relax-and-cut MWCS solver (Alvarez-Miranda and Sinnl, 2017):

```{r}
rcsolver <- rmwcs_solver()
```

Now we can use this solver to solve the example instance:

```{r}
m <- solve_mwcsp(rcsolver, mwcs_example)
print(m$graph)
print(m$weight)
```

## Using exact CPLEX-based Virgo solver

The `mwcsr` package also provide and interface to exact CPLEX-based Virgo solver  
(https://github.com/ctlab/virgo-solver) which can be used to solve
MWCS, GMWCS and SGMWCS instances to provable optimality.

To setup this solver CPLEX libraries has to be available. 
CPLEX can be downloaded from the official web-site:
https://www.ibm.com/products/ilog-cplex-optimization-studio.
Free licence can be obtained for academic purposes.
    
First, initialize `cplex_dir` variable to contain path to CPLEX libraries (for example, `r params$cplex_dir`). 

    
```{r eval=FALSE}
cplex_dir <- '<replace with path to cplex folder>'
```

```{r echo=FALSE}
cplex_dir <- params$cplex_dir
```

Then initialize the solver:

```{r}
vsolver <- virgo_solver(cplex_dir=cplex_dir)
```

And run it on the same MWCS instance:

```{r}
m <- solve_mwcsp(vsolver, mwcs_example)
print(m$graph)
```

While the solution is a bit different its weight is the same as found before.
The solutions differs only in zero-weight vertices.

```{r}
get_weight(m$graph)
```

However, Virgo guarantees that the result is optimal, unless the solver was interrupted 
on time limit.

```{r}
m$solved_to_optimality
```


Next, consider a GMWCS instance which additionally has edge weights:

```{r}
data("gmwcs_example")
gmwcs_example
summary(E(gmwcs_example)$weight)
```

The same solver can be used to solve this instance:

```{r}
m <- solve_mwcsp(vsolver, gmwcs_example)
print(m$graph)
get_weight(m$graph)
```

Finally, let consider an SGMWCS instance. The weights of nodes and edges are defined not 
directly, but through the `signals` attribute:

```{r}
data("sgmwcs_example")
sgmwcs_example
str(V(sgmwcs_example)$signal)
str(E(sgmwcs_example)$signal)
head(sgmwcs_example$signals)
```

Again, we can solve this instance with Virgo solver:

```{r}
m <- solve_mwcsp(vsolver, sgmwcs_example)
print(m$graph)
get_weight(m$graph)
```

## Running Virgo heuristics without CPLEX

In case CPLEX is not available, Virgo solver can be run in the heuristic mode. 
Just set `cplex_dir` parameter to `NULL`:

```{r}
vhsolver <- virgo_solver(cplex_dir=NULL)
```

While the results are not optimal, sometimes they can be enough for practical applications:

```{r}
m <- solve_mwcsp(vhsolver, mwcs_example)
get_weight(m$graph)
m$solved_to_optimality

m <- solve_mwcsp(vhsolver, gmwcs_example)
get_weight(m$graph)

m <- solve_mwcsp(vhsolver, sgmwcs_example)
get_weight(m$graph)
```


